---
title: "The Data-Type Ontology"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The Data-Type Ontology}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = FALSE, collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
library(formex)
o <- fx_ontology()
```

## Why a column needs more than a type

A CSV cell like `47401` is just text. Is it a ZIP code, a county FIPS, a count
of something? A single "type" label cannot capture that: `47401` and `06037` are
both five-digit geographic codes (same *type*), but one is a ZIP and one is a
county (different *meaning*), and neither should be summed like a number
(different *treatment*). formex pulls these apart onto a small set of
coordinates, held in the **research data-type ontology** --- the shared
vocabulary that [`guess_data_type()`](../reference/guess_data_type.html) and
[`fx_stabilize()`](../reference/fx_stabilize.html) both speak.

Read the ontology with `fx_ontology()`; it ships as a diff-able CSV
(`inst/extdata/research_data_type_ontology_v6.csv`) and a parsed `data/ontology.rda`,
and a package test asserts the two agree.

```{r ontology-shape, echo = TRUE}
o <- fx_ontology()
dim(o)
```

Each **row is one format variant** of a semantic type. The columns fall into
four groups: the **path** (what a column *is*), the **detection kind** (how it
is recognized), the **recipe** (how raw values are stabilized), and
**metadata** (units, validation, lifecycle).

## The path: four coordinates

A column is placed on a four-level path, coarse to fine:

```
data_type  ›  semantic_family  ›  semantic_type  ›  format_label
 treatment      broad role         specific meaning     surface form
```

* **`data_type`** --- how a value is *treated*: `number`, `text`, `boolean`,
  `temporal`, `categorical`, `identifier`, `structured`, `unknown`.
* **`semantic_family`** --- the broad role within a type (e.g. `currency`,
  `name`, `date`, `geographic`).
* **`semantic_type`** --- the specific meaning that carries the validation and
  the render treatment (e.g. `usd`, `person`, `calendar`, `us_state`).
* **`format_label`** --- the surface *form* the file happens to use for that
  meaning; one semantic type can have several (e.g. an ISBN written 10-digit,
  13-digit, or hyphenated). The `format_label` selects the recipe row.

`data_type` through `semantic_type` say what a column **is** --- its intended
meaning, stable across file versions. `format_label` describes how the *current*
file writes it. That is why `guess_data_type()` can return a **path prefix** when
it is confident about the type but not the exact format: it stops at the deepest
coordinate the values actually support.

Every semantic type also carries a stable `semantic_type_id` (e.g. `n0059`), and
each variant a `variant_id` (e.g. `n0059-f01`). Ids are allocated once and never
renumbered, so banked test cases and downstream governance files stay valid
across ontology versions.

### The full hierarchy

Built live from `fx_ontology()` --- `data_type › semantic_family : semantic_types`:

```{r hierarchy, results = "asis"}
h <- unique(o[order(o$data_type, o$semantic_family, o$semantic_type),
              c("data_type", "semantic_family", "semantic_type")])
for (dt in unique(h$data_type)) {
  cat("\n**`", dt, "`**\n\n", sep = "")
  fam <- h[h$data_type == dt, ]
  for (fm in unique(fam$semantic_family)) {
    sts <- fam$semantic_type[fam$semantic_family == fm]
    cat("* `", fm, "` --- ", paste0("`", sts, "`", collapse = ", "), "\n", sep = "")
  }
}
```

## The detection kind

`match_kind` records *how* a semantic type is recognized from values --- which
is also what decides whether a column can be typed from its contents at all.

```{r match-kind}
knitr::kable(as.data.frame(table(match_kind = o$match_kind),
                           responseName = "rows"))
```

| `match_kind` | recognized by | examples |
|---|---|---|
| **`genome`** | a tolerant multi-format parser | `calendar`, `timestamp`, `clock` |
| **`mask`** | a regex / format mask (often checksum-validated) | `ein`, `isbn`, `phone`, `usd` |
| **`lexicon`** | membership in a controlled vocabulary | `us_state`, `iso_country`, `first_name` |
| **`mask+lexicon`** | a shape gate, then a vocabulary check | `person` (name-shaped *and* in the gazetteer) |
| **`shape`** | a numeric distribution | `count`, `ratio`, `probability` |
| **`none`** | not value-detectable; only metadata can assert it | `record` keys, free `text`, `unknown` |

`none`-kind types are the reason some columns can never be pinned from values
alone: a surrogate record key is just an integer, and free text is just text.
Those are asserted by context (a column name, a data dictionary), not by
`guess_data_type()`.

## The recipe: raw → stable → import

Four columns encode how a raw column is turned into a canonical, analysis-ready
form. They use the **formex** expression syntax (`{{ ... }}`); the *Formex*
vignette is the full guide.

| column | holds | example |
|---|---|---|
| **`data_format`** | the mask the *raw* file matches | `{{ mmddyyyy }}`, `{{ [strip:$,][dec:.2] }}` |
| **`raw_to_stable_transform`** | a *lossless* rewrite to canonical form | `{{ as_mmddyyyy }}`, `{{ as_usd }}`, `lookup:usps` |
| **`stable_format`** | what a stabilized value looks like | `2019-03-15`, `CA` |
| **`stable_import_rule`** | how to read the stable column into its type | `{{ as_yyyymmdd }}`, `{{ as_integer }}` |

A few real rows, straight from the ontology:

```{r recipe-table}
# unique semantic types spanning the match kinds (genome / mask / mask+lexicon / shape)
show <- c("calendar", "administrative", "usd", "person", "count")
r <- do.call(rbind, lapply(show, function(st) o[o$semantic_type == st, ][1, ]))
knitr::kable(r[, c("semantic_type", "format_label", "match_kind",
                   "data_format", "raw_to_stable_transform", "stable_format")],
             row.names = FALSE)
```

The split has a purpose: push as much as possible into the *lossless* reformat
(`raw_to_stable_transform`), which every downstream tool benefits from, and
reserve `stable_import_rule` for the genuinely schema-dependent step of coercing
a value into a type a plain CSV reader cannot infer (a ZIP that must stay a
`categorical`, an EIN that must stay an `identifier`).

## Metadata: units, validation, lifecycle

The remaining columns are documentation the engine carries but does not parse:

* **`dd_data_unit`** --- an interpretive unit an analyst needs but a type does
  not (`USD`, `kg`, `EPSG:4326`). It is proposed only when the semantic type
  declares one. Crucially it does **not** carry temporal *grain*: a date's grain
  (calendar vs month vs day-of-week) selects the render graphic, so it lives in
  `semantic_type`, not in a unit.
* **`validation`**, **`default_storage`**, **`short_desc`** / **`detail`** ---
  the human-facing description, the storage class a stabilized value round-trips
  to, and the checks a value should pass.
* **`implemented`** --- whether a value-level detector/recipe is wired up yet, so
  the ontology can name a type before its machinery exists.
* **`version_added`** / **`version_deprecated`** / **`superseded_by`** ---
  lifecycle. Rows are deprecated, never deleted, so old labels remain resolvable.

## How the engine uses the ontology

**`guess_data_type()`** anchors a column to these coordinates. The `$route`
record is the ontology path it resolved to, only as deep as it is confident:

```{r guess, echo = TRUE}
r <- guess_data_type(c("03/15/2019", "12/31/2020", "01/01/2021"))$route
r[c("data_type", "semantic_family", "semantic_type", "format_label", "match_kind", "depth")]
```

**`fx_stabilize()`** looks up the recipe for a `(semantic_type, format_label)`
pair and runs it, returning the stable values plus the format mask and import
rule from that row:

```{r stabilize, echo = TRUE}
fx_stabilize(c("California", "TX", "Ohio"), "geographic", "us_state")[
  c("stable", "stable_format", "import_rule")]
```

Because both functions read the *same* `fx_ontology()`, a column typed by
`guess_data_type()` can be handed straight to `fx_stabilize()` using the
`semantic_type` and `format_label` it returned.

## See also

* The **Guessing a Column's Type** vignette --- the router → verifier flow that
  produces the `$route` path above.
* The **Data-Type Detector Guide** --- the `is_*` predicates behind the
  `mask` / `lexicon` / `genome` match kinds.
* `fx_ontology()`, `fx_validate_ontology()`, and
  `inst/extdata/research_data_type_ontology_v6.csv` --- the accessor, the
  integrity check, and the full diff-able catalog.
```
