---
title: "Guessing a Column's Type"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Guessing a Column's Type}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, collapse = TRUE, comment = "#>",
                      message = FALSE, warning = FALSE)
library(formex)
```

## The chooser above the detectors

The *Data-Type Detector Guide* describes the detector library: dozens of pure
predicates that each answer one yes/no question about a column
(`is_currency_code()`, `is_iso_date()`, `is_county_fips()`, ...).
`guess_data_type()` sits **above** them. Given a column of values it returns two
aligned views: a v6 ontology **`$route`** --- the current two-stage
router → verifier answer --- and a set of **legacy** detector fields kept for
backward compatibility.

## The `$route`: an ontology path, only as deep as it's sure

`$route` anchors the column to the ontology
(`data_type › semantic_family › semantic_type › format_label`; see the
*Data-Type Ontology* vignette) and stops at the deepest coordinate the values
actually support:

```{r route}
r <- guess_data_type(c("03/15/2019", "12/31/2020", "01/01/2021"))$route
r[c("data_type", "semantic_family", "semantic_type", "format_label",
    "match_kind", "depth", "confidence", "router_method")]
```

* **`depth`** records how far it resolved: `0` unknown · `1` data_type ·
  `2` family · `3` semantic_type · `4` leaf (format). Fields below the confident
  depth are `NA`. A column of clean dates resolves all the way to a leaf; a
  column of ambiguous five-digit codes may stop at its data_type. This
  **graceful truncation** is deliberate --- forcing a leaf would manufacture
  false precision.
* **`match_kind`** is how the resolved type is recognized (genome / mask /
  lexicon / ...).
* **`router_method`** reports which router produced the shortlist (below).

### Two stages: router, then verifier

* **Router** (stage 1) shortlists candidate semantic types. Two ship: a
  **learned** xgboost model over the full value profile
  (`fx_extract_features()`), and a model-free **signature** router. The
  `router =` argument chooses (`"auto"`, `"learned"`, `"signature"`); with
  `xgboost` absent, `"auto"` falls back to the signature router.
* **Verifier** (stage 2) confirms a leaf deterministically with
  *confirm = detect ∘ transform*, gated so that permissive types truncate
  instead of claiming a false leaf.

```{r router-arg}
guess_data_type(c("14:30:00", "09:15:45", "23:59:59"),
                router = "signature")$route[c("semantic_type", "depth")]
```

## The legacy detector view

The same call also returns the flat detector fields it always did, so existing
code keeps working:

```{r legacy}
g <- guess_data_type(c("USD", "EUR", "JPY", "GBP"))
g$guess
g$ontology
```

* **`guess`** --- the winning detector name, or `NA` when nothing is confident.
* **`ontology`** --- the flat `c(data_type, data_subtype, data_class,
  data_format)` coordinates for that detector (the legacy scheme; the v6 path is
  in `$route`).
* **`confidence`** --- the winner's pass rate (share of values it matched).
* **`candidates`** --- every type that cleared the reporting `floor`, ranked.
* **`n`** --- how many unique values were actually scored.

The legacy path stays cheap on big data: it **dedupes and samples** (only unique
non-missing values, at most `n_sample`, default 100), scores them against every
registered detector, and ranks by pass rate. The raw scoring step is exposed on
its own as `score_data_type()`:

```{r score}
score_data_type(c("USD", "EUR", "XYZ"))[1, ]
```

## The name hint breaks ties the data can't

Some values match more than one type. A five-digit code like `06037` is a valid
ZIP **and** a valid county FIPS --- both detectors pass at 100%. The data alone
cannot choose; the **variable name** can (it is a tie-break for the legacy
`$guess`):

```{r namehint}
vals <- c("06037", "36061", "48201", "17031")

guess_data_type(vals, name = "county_fips")$guess
guess_data_type(vals, name = "zip_code")$guess
```

Look at the ranked candidates to see *why*: both types pass at `1.0`, so the
`name_match` column casts the deciding vote, and with no name the tie falls to
the more **specific** detector.

```{r candidates}
guess_data_type(vals, name = "county_fips")$candidates
```

## Confidence, and the `threshold` / `floor` knobs

When nothing matches well, the honest answer is "I don't know":

```{r unknown}
guess_data_type(c("apple", "banana", "cherry"))$guess
```

Two thresholds control how bold the legacy guess is:

* **`threshold`** (default `0.8`) --- the top type must match at least this share
  of values to be reported as `guess`; below it, `guess` is `NA`.
* **`floor`** (default `0.5`) --- the minimum pass rate to appear in `candidates`.

```{r knobs}
# a column that is 3/4 ISO dates: unknown at the default, a date if you relax it
messy <- c("2016-03-15", "2016-03-13", "1999-12-31", "not a date")
guess_data_type(messy)$guess               # NA (below 0.8)
guess_data_type(messy, threshold = 0.7)$guess
```

## From a guess to a stable column

Because `$route` returns the ontology `semantic_type` and `format_label`, its
output hands straight to `fx_stabilize()` to normalize the column:

```{r stabilize}
x     <- c("03/15/2019", "12/31/2020", "01/01/2021")
route <- guess_data_type(x)$route
fx_stabilize(x, route$semantic_type, route$format_label)$stable
```

That is the whole formex loop: `guess_data_type()` says *what a column is*, and
`fx_stabilize()` puts it in a standard form. A downstream workflow package (such
as datagoodr) can build a per-column governance file on top of these two calls
--- see the *Data-Type Ontology* vignette for the coordinates they share.
